---
output:
  html_document:
    keep_md: yes
---
Activity Monitoring Evaluation
========================================================================================

Loading and pre-processing the data

```{r}
setwd("~/.")
library(dplyr)
library(lubridate)
library(xtable)
library(ggplot2)
if(!file.exists("./ActivityData")){
    file_url <- ("https://github.com/ALCrist/RepData_PeerAssessment1/blob/master/activity.zip?raw=true
")    
    dir.create("ActivityData", recursive=FALSE,showWarnings=TRUE)
    download.file(file_url, destfile = "./ActivityData/activity.zip", mode="wb")
    
}
path <- path.expand("~/ActivityData/activity.zip")
activity <- read.csv(unzip(path),
                       header = TRUE, stringsAsFactors=FALSE)
```
####  What is mean total number of steps taken per day?

Calculate the total number of steps taken per day:
```{r TotalSteps}
TotalSteps <- summarize(group_by(activity,date),steps=sum(steps))
```

This is a histogram of the total number of steps taken each day:

```{r HistTotalSteps}
par(mar=c(4,4,2,1))
hist(TotalSteps$steps, xlab="Steps", main="Total Steps Taken Per Day", col="blue")

```




Calculate the mean and median number of steps per day:
```{r AvgSteps}
MeanSteps <- as.data.frame(summarize(group_by(activity,Date=date), Mean=mean(steps), 
                       Median=median(steps)))
```

This is the mean and median number of steps taken each day:
```{r TabAvgSteps, results='asis', fig.align='center'}
ms <- xtable(MeanSteps[which(!is.na(MeanSteps$Mean)),], digits=2)
print(ms, type="html")

```
####  What is the average daily activity pattern?

Calculate the mean number of steps per interval over all days:
```{r DailyActivity}
DailyActivity <- as.data.frame(summarize(group_by(activity, Interval=interval),Mean=mean(steps, na.rm=TRUE)))                           
```

This is a graph of the average daily activity:
```{r DailyActivityPlot}

par(mar=c(4,4,2,1))
plot(DailyActivity$Interval,DailyActivity$Mean, type="l", col="blue",
     main="Average Daily Activity Pattern",xlab="Interval",ylab="Avg. Steps")
```

Determine interval with maximum average number of steps:

```{r MaxInterval}
maxint <- DailyActivity[which(DailyActivity$Mean==max(DailyActivity$Mean)),"Interval"]

```

Interval **`r maxint`** on average has the maximum number of steps.


####  Imputing missing values

Replace missing steps value with mean steps for given interval.

```{r NewActivitySet}
na.activity <- activity[which(is.na(activity$steps)),]
merged.na.activity <- merge(na.activity, DailyActivity, by.x="interval", by.y="Interval")
merged.na.activity$steps <- merged.na.activity$Mean
new.activity <- rbind(merged.na.activity[,c("interval","steps","date")],
                      activity[which(!is.na(activity$steps)),])
```

**Imputing the mean steps for a given interval increased the frequency of 10,000 to 15,000 total steps taken by 8.**  


This is a histogram of the total number of steps taken each day using updated data set:

```{r HistNewTotalSteps}
newTotalSteps <- summarize(group_by(new.activity,date),steps=sum(steps))

par(mar=c(4,4,2,1))
hist(newTotalSteps$steps, xlab="Steps", main="Total Steps Taken Per Day", col="blue")
```

**The median number of steps for days with imputed values is significantly higher than the median number of steps for those days with actual measures.**

This is the mean and median number of steps taken each day using updated data set:
```{r NewTabAvgSteps, results='asis', fig.align='center'}

newMeanSteps <- as.data.frame(summarize(group_by(new.activity,Date=date), Mean=mean(steps), 
                                     Median=median(steps)))
newms <- xtable(newMeanSteps, digits=2)
print(newms, type="html", methods=print.xtable)
```

####  Are there differences in activity parterns between weekdays and weekends?

Calculate the mean number of steps for all weekday and weekend days:
```{r }
new.activity <- cbind(new.activity, day=wday(ymd(new.activity$date)))
new.activity$factor <- as.factor(ifelse(new.activity$day %in% c(2:6),"Weekday","Weekend"))

new.DailyActivity <- as.data.frame(summarize(group_by(new.activity, Interval=interval,DayType=factor),Mean=mean(steps, na.rm=TRUE)))                           
```

This is a graph of the average daily activity split between weekday and weekend days:
```{r DailyActivityComp }
p2 <-ggplot(new.DailyActivity, aes(Interval, Mean)) 
p2 <- p2 + geom_line(col="blue") + facet_grid(DayType~.) +
    xlab("Intervals") + ylab("Number of Steps") +
    theme_bw() + labs(title = "Average Daily Activity") 
print(p2)
```
